---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
library(ggplot2)
library(ggmap)
```

# PatricksPlants

<!-- badges: start -->
<!-- badges: end -->

The goal of PatricksPlants is to make Patrick Alexander's plant occurance observations more accessible for analysis. 

A data frame is provided with all observations and metadata coded as presence-only. Presence-absence data may reasonably be infered from this and a function will be provided to do so for a specified set of species or for the entire dataset. Beware that presence-absence can get rather large for cross-product of sites and plants in the dataset. At the time of this writing, the presence-only data were around 10,000 rows, while the presence-absence data were over 10,000,000 rows.

A function is provided for reading Patrick's xlsx file, in which he records his obsevations from his field notebook. The function returns a data frame of presence-only or presence-absence with metadata for sites, surveys, and plant taxonomy. The presence-only output is identical in format to the supplied dataset.

## Installation

<!-- You can install the released version of PatricksPlants from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("PatricksPlants") -->
<!-- ``` -->

You ~~can~~ _will be able to_ install the development version of PatricksPlants from GitHub with:

``` r
install.packages("devtools")
devtools::install_github("gregorypenn/PatricksPlants")
```

## Examples

Here are the locations in the dataset.

```{r example, message=FALSE}
library(PatricksPlants)
library(ggplot2)

qmplot(longitude, latitude, 
       data = patricks_plants)
```

What's the distribution of species counts across surveys?
```{r}
species_richness <- patricks_plants %>%
  filter(area_approx == "10m radius") %>%
  group_by(survey_id) %>%
  summarize(plants = n())

ggplot(species_richness, aes(x = plants)) +
  geom_histogram(binwidth = 1) +
  theme_bw()

```

How many survey sites had *Bouteloua eriopoda*, and how many didn't?
```{r}
BOER4 <- presence_absence("BOER4", patricks_plants)
table(BOER4$presence)
```
<!-- Don't forget to commit and push figure files, so they display on GitHub! -->
